Synthesis of Linear Functional Programs

### 2021-05-09 article/article.tex
.

\mypara{Propositions as Types}
%
It turns out that the inference rules of the simply-typed
$\lambda$-calculus are closely related to those of a system of natural
deduction for intuitionistic logic~\cite{prawitznd65}. This
relationship, known as the Curry-Howard correspondence~\cite{},
identifies that the propositions of intuitionistic logic can be read
as types for the simply-typed $\lambda$-calculus (``propositions as types'' observation),
their proofs are exactly the program with the given type (``proofs as programs'' observation),
and checking a proof is type checking a program (``proof checking as type checking'' observation)~\cite{}.

Inference rules in natural deduction are categorized as introduction or elimination, these
correspond to constructors and destructors in programming languages
(e.g. function abstraction vs application, construction of a pair vs
projection).

We introduce a few typing rules to both show how a type system can be
formalized and to show the relationship with (intuitionistic)
propositional logic. The following rule captures the nature of a
hypothetical judgment, allowing for reasoning from assumptions:
\[
    \infer*[right=(var)]
    {  }
    {\Gamma, u: A \vdash u : A}
\]
When seen as a typing rule for the $\lambda$-calculus, it
corresponds to the rule for typing variables -- variable $u$ has
type $A$ if the typing environment contains a variable $u$ of
type $A$.

@@

Finally, let us consider an \emph{elimination} rule, also for the implication connective:
\[
    \infer*[right=($\rightarrow E$)]
    {\Gamma \vdash M : A \rightarrow B \and \Gamma \vdash N : A}
    {\Gamma \vdash M\,N : B}
\]
Elimination rules are easier to think of in a top-down manner. Logically, this rule states
that if we prove $A \rightarrow B$ and $A$ we can prove $B$. Through the Curry-Howard correspondence,
implication elimination corresponds to function application, and its type is the function return type
-- the program $M N$ has type $B$, provided $M$ has type $A \rightarrow B$ and $N$ has type $A$.

The Curry-Howard correspondence generalizes beyond the simply-typed
$\lambda$-calculus and propositional intuitionistic logic. It extends to
the realms of polymorphism (higher-order logic \todo{review - {https://en.wikipedia.org/wiki/Lambda_cube}} second-order propositional calculus),
dependent types (first-order predicate logic), and every other extension
of natural deduction and simply-typed $\lambda$-calculus -- which also means
linear logic and linear types.

\todo[inline]{Daqui, para tipificar uma função, para Curry-Howard,
  para lógica linear, para proof search, etc\dots Só faz sentido falar de dedução natural no
  contexto de ``lógica'' / Curry-Howard.}

@@

To present the formalization of this logic, besides the new
connectives, we need to introduce the \emph{resource-aware context}
$\Delta$.  In contrast to the previously seen $\Gamma$, $\Delta$ is
also a list of variables and their types, but where each and every
variable must be used exactly once during inference.  So, to introduce
the connective $\tensor$ which defines a multiplicative pair of
propositions, we must use exactly all the resources
($\Delta_1, \Delta_2$) needed to realize the ($\Delta_1$) 
proposition $A$, and ($\Delta_2$) proposition $B$:
\[
    \infer*[right=($\tensor I$)]
    {\Delta_1 \vdash M : A \and \Delta_2 \vdash N : B}
    {\Delta_1 , \Delta_2 \vdash (M \tensor N) : A \tensor B}
\]
Out of the logical connectives, we need to mention one more, since
it'll alter the judgment form, and it's the one that ensures logical
strength -- i.e. we're able to translate intuitionistic logic into
linear logic.
  The proposition $\bang A$ (read \emph{of course} $A$)
is used (under certain conditions) to make a resource ``infinite''
i.e. to make it useable an arbitrary number of times. To distinguish the
``infinite'' variables, a separate, unrestricted, context is used -- $\Gamma$. So
$\Gamma$ holds the ``infinite'' resources, and $\Delta$ the resources
that can only be used once.
The linear typing judgment for the introduction of the exponential $\bang$ takes the form:
\todo{A formula está um pouco estranha (faltam espaços)?}
\[
    \infer*[right=($\bang I$)]
    {\Gamma ; \emptyset \vdash M : A}
    {\Gamma ; \emptyset \vdash !M : !A}
\]
\todo{Describe exponent introduction and new gamma context}
\todo{Add h spaces in formula}
\[
    \infer*[right=($\bang E$)]
    {\Gamma ; \Delta_1 \vdash M : !A \and \Gamma, u{:}A ; \Delta_2 \vdash N : C}
    {\Gamma ; \Delta_1, \Delta_2 \vdash let !u = M \mathsf{in} N : C}
\]
\todo{Describe exponent elimination and new gamma context}

@@

Type-based program synthesis is a studied field\todo{ dizer de uma forma melhor? }, and it follows
that a lot of literature is available to inspire and complement our work.
Most? follow the synthesis-as-proof-search pattern, however, the process is
novel for each, due to a variety of different rich types explored, and their respective logics; or nuances done to the process, such
as considering examples; or even the programming paradigm of the output produced (e.g. generating heap manipulating programs~\cite{DBLP:journals/pacmpl/PolikarpovaS19}).
Some other common patterns we might want to follow are the type refinements and the support for recursive functions.
These works share some common ground, and base themselves of the same works we do -- most famously the Curry-Howard correspondence, but also,
for example, works on the method of Focusing~\cite{focusing}.

\mypara{Type-and-Example-Directed Program Synthesis}~\cite{DBLP:conf/pldi/OseraZ15,DBLP:conf/popl/FrankleOWZ16}
This paper explores a purely functional, recursive program synthesis technique based on types/proof search (as we intend to do),
with examples as an auxiliar technique to trim down the program synthesis vast search space.
\todo[inline]{more introduction to the paper}.
A good use is made of the general strategy of using a type system
to generate programs (by ``inverting'' the type system), rather than using it to type-check one.
Moreover, the paper uses a data structure called ``refinement tree'' in conjunction
with the extra information provided by the examples to allow for efficient synthesis
of non-trivial programs. Overall it employs good engineering to achieve type-based synthesis
of functional typed programs, with at least two additions that we might want to follow too
(type refinements and recursive functions).

>>>>

\mypara{Propositions as Types}
%
It turns out that the inference rules of the simply-typed
$\lambda$-calculus are closely related to those of a system of natural
deduction for intuitionistic logic~\cite{prawitznd65}. This
relationship, known as the Curry-Howard correspondence~\cite{},
identifies that the propositions of intuitionistic logic can be read
as types for the simply-typed $\lambda$-calculus (``propositions as
types''), their proofs are exactly the program with the given type
(``proofs as programs''), and checking a proof is type checking a
program (``proof checking as type checking'')~\cite{}.

Inference rules in natural deduction are categorized as introduction
or elimination rules, these correspond, respectively, to rules for
constructors and destructors in programming languages (e.g. function
abstraction vs application, construction of a pair vs projection).

We introduce select typing rules to both show how a type system can be
formalized and to show the relationship with (intuitionistic)
propositional logic. The following rule captures the nature of a
hypothetical judgment, allowing for reasoning from assumptions:
\[
    \infer*[right=(var)]
    {  }
    {\Gamma, u: A \vdash u : A}
\]
When seen as a typing rule for the $\lambda$-calculus, it
corresponds to the rule for typing variables -- variable $u$ has
type $A$ if the typing environment contains a variable $u$ of
type $A$.

@@

Finally, let us consider an \emph{elimination} rule, also for implication:
\[
    \infer*[right=($\rightarrow E$)]
    {\Gamma \vdash M : A \rightarrow B \and \Gamma \vdash N : A}
    {\Gamma \vdash M\,N : B}
\]
Elimination rules are easier to think of in a top-down
manner. Logically, this rule states that if we prove $A \rightarrow B$
and $A$ we can prove $B$. Through the Curry-Howard correspondence,
implication elimination corresponds to function application, and its
type is the function return type -- the application $M N$ has type
$B$, provided $M$ has type $A \rightarrow B$ and $N$ has type $A$.

The Curry-Howard correspondence generalizes beyond the simply-typed
$\lambda$-calculus and propositional intuitionistic logic. It extends
to the realms of polymorphism (higher-order logic \todo{review -
  \url{https://en.wikipedia.org/wiki/Lambda_cube}} second-order
propositional calculus), dependent types (first-order predicate
logic), and various other extensions of natural deduction and
simply-typed $\lambda$-calculus -- which include linear logic and
linear types.

@@

To present the formalization of this logic, besides the new
connectives, we need to introduce the \emph{resource-aware context}
$\Delta$.  In contrast to the previously seen $\Gamma$, $\Delta$ is
also a list of variables and their types, but where each and every
variable must be used exactly once during inference.  So, to introduce
the connective $\tensor$ which defines a multiplicative pair of
propositions, we must use exactly all the resources
($\Delta_1, \Delta_2$) needed to realize the ($\Delta_1$) 
proposition $A$, and ($\Delta_2$) proposition $B$:
\[
    \infer*[right=($\tensor I$)]
    {\Delta_1 \vdash M : A \and \Delta_2 \vdash N : B}
    {\Delta_1 , \Delta_2 \vdash (M \tensor N) : A \tensor B}
\]
Out of the logical connectives, we need to mention one more, since it
augments the form of the judgment and it's the one that ensures
logical strength -- i.e. we're able to translate intuitionistic logic
into linear logic.  The proposition $\bang A$ (read \emph{of course}
$A$) is used (under certain conditions) to make a resource
``infinite'' i.e. to make it useable an arbitrary number of times. To
distinguish the ``infinite'' variables, a separate, unrestricted,
context is used -- $\Gamma$. So $\Gamma$ holds the ``infinite''
resources, and $\Delta$ the resources that can only be used once.  The
linear typing judgment for the introduction of the exponential $\bang A$
takes the form: 
\[
    \infer*[right=($\bang I$)]
    {\Gamma ; \emptyset \vdash M : A}
    {\Gamma ; \emptyset \vdash \bang M : \bang A}
  \]
  Logically, a proof of $\bang A$ cannot use linear resources since
  $\bang A$ denotes an unbounded (potentially $0$) number of copies of
  $A$. Proofs of $\bang A$ may use other unrestricted or exponential
  resources, tracked by context $\Gamma$.
  From a computational perspective, the type $\bang A$
  internalizes the simply-typed $\lambda$-calculus in the linear
  $\lambda$-calculus.

  The elimination form for the exponential, written $\vdash \llet{ !u
    = M}{ N}$, warrants the use of the resource $A$ an unbounded
  number of times in $N$ via the variable $u$:  
\[
    \infer*[right=($\bang E$)]
    {\Gamma ; \Delta_1 \vdash M : \bang A \and \Gamma, u{:}A ; \Delta_2 \vdash N : C}
    {\Gamma ; \Delta_1, \Delta_2 \vdash \llet{ !u = M}{ N} : C}
\]

@@

Type-based program synthesis is a vast field of study and so it
follows that a lot of literature is available to inspire and
complement our work. Most works~\cite{} follow some variation of the
synthesis-as-proof-search approach, however, the process is novel for
each, due to a variety of different rich types explored, and their
respective corresponding logics and programming languages; or nuances
of the synthesis process itself, such as complementing types with
program examples; or even the programming paradigm of the output
produced (e.g. generating heap manipulating
programs~\cite{DBLP:journals/pacmpl/PolikarpovaS19}).  Some other
common patterns we might want to follow are the use of type
refinements~\cite{} and the support for synthesis of recursive
functions~\cite{}.  These works share some common ground, and base
themselves of the same works we do -- most famously the Curry-Howard
correspondence, but also, for example, works on focusing-based
proof-search~\cite{focusing}.

\mypara{Type-and-Example-Directed Program Synthesis} This
work~\cite{DBLP:conf/pldi/OseraZ15,DBLP:conf/popl/FrankleOWZ16}
explores a purely functional, recursive program synthesis technique
based on types/proof search (as we intend to do), with examples as an
auxiliar technique to trim down the program synthesis vast search
space.  \todo[inline]{more introduction to the paper}.  A good use is
made of the general strategy of using a type system to generate
programs (by ``inverting'' the type system), rather than using it to
type-check one.  Moreover, the paper uses a data structure called
``refinement tree'' in conjunction with the extra information provided
by the examples to allow for efficient synthesis of non-trivial
programs. Overall it employs good engineering to achieve type-based
synthesis of functional typed programs, with at least two additions
that we might want to follow too (type refinements and recursive
functions).
